---
title: "Factor Model Analysis of Vietnamese Stock Market 2023 (HOSE)"
author: "Minh Duc Nguyen"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

# 1. Library

```{r libraries}
library(readxl)
library(lmtest)
library(sandwich)
library(car)
library(zoo)
library(ggplot2)
library(corrplot)
library(knitr)
library(dplyr)

PATH_PRICE <- "data/dinhgia1.50_cp_hose.xlsx"   
PATH_META  <- "data/dinhgia1.vh.xlsx"     

RISK_FREE_ANNUAL <- 0.0321          
N_DAYS           <- 248             
N_STOCKS         <- 50          
```


```{r load-prices}
raw <- read_excel(PATH_PRICE)
price_data <- raw[, -1]   
```

# 2. Log-return
```{r calc-returns}
# log-return: r_t = ln(P_t / P_{t-1}) ---
calc_log_return <- function(price_vec) {
  p <- as.numeric(price_vec)
  diff(log(p))                  
}

r <- as.data.frame(
  lapply(price_data, calc_log_return)
)
colnames(r) <- colnames(price_data)
```


# 3. Fama-French Factor

## 3.1. F1 — Market Factor (Mkt-RF)

```{r factor-f1}
# Log-return of VNINDEX
r_VnIndex <- r[["VnIndex"]]

# Daily free-risk rate
r_f <- rep((1 + RISK_FREE_ANNUAL)^(1 / 365) - 1, nrow(r))

# F1 = Excess market return
F1 <- r_VnIndex - r_f

```

## 3.2. F2 — SMB Factor (Small Minus Big)

```{r factor-f2}
meta  <- read_excel(PATH_META)
#Capitalization
CSVH  <- as.numeric(meta[1, -1])      

med_cap <- median(CSVH)
idx_big   <- which(CSVH > med_cap)        
idx_small <- which(CSVH < med_cap)         

r_big   <- r[, idx_big,   drop = FALSE] #Big capitalization stock
r_small <- r[, idx_small, drop = FALSE] #Small capitalization stock

# F2 = Average(Small) - Average(Big)
F2 <- rowMeans(r_small) - rowMeans(r_big)
```

## 3.3. F3 — HML Factor (High Minus Low BtM)

```{r factor-f3}
# Book-to-Market ratio
BtM <- as.numeric(meta[2, -1])

q33 <- quantile(BtM, 1/3, na.rm = TRUE)
q67 <- quantile(BtM, 2/3, na.rm = TRUE)

idx_low  <- which(BtM < q33)              # Low BtM 
idx_high <- which(BtM > q67)              # High BtM

r_high <- r[, idx_high, drop = FALSE]
r_low  <- r[, idx_low,  drop = FALSE]

# F3 = Average(High BtM) - Average(Low BtM)
F3 <- rowMeans(r_high) - rowMeans(r_low)

```

# 4. Correlation Between Factors

```{r factor-corr}
factor_mat <- data.frame(F1_Market = F1, F2_SMB = F2, F3_HML = F3)
corr_mat   <- cor(factor_mat)

kable(round(corr_mat, 4), caption = "Correlation Matrix Between 3 Factors")

corrplot(corr_mat,
         method   = "color",
         type     = "upper",
         addCoef.col = "black",
         tl.col   = "black",
         title    = "Correlation Between Fama-French Factors",
         mar      = c(0, 0, 2, 0))
```

# 5. Fama-French Regression

```{r reg-all-stocks}
models     <- vector("list", N_STOCKS)
coef_table <- data.frame(matrix(NA, N_STOCKS, 4))
colnames(coef_table) <- c("alpha", "Beta1", "Beta2", "Beta3")

for (i in 1:N_STOCKS) {
  models[[i]]     <- lm(r[, i] ~ F1 + F2 + F3)
  coef_table[i, ] <- coef(models[[i]])
}
rownames(coef_table) <- colnames(r)[1:N_STOCKS]
```

```{r plot-betas}
# Distribution of Beta coefficients across 50 stocks
par(mfrow = c(1, 3))
hist(coef_table$Beta1, main = "Distribution of Beta1 (Market)",
     xlab = "Beta1", col = "#4E79A7", border = "white", breaks = 15)
abline(v = mean(coef_table$Beta1), col = "red", lwd = 2, lty = 2)

hist(coef_table$Beta2, main = "Distribution of Beta2 (SMB)",
     xlab = "Beta2", col = "#F28E2B", border = "white", breaks = 15)
abline(v = mean(coef_table$Beta2), col = "red", lwd = 2, lty = 2)

hist(coef_table$Beta3, main = "Distribution of Beta3 (HML)",
     xlab = "Beta3", col = "#59A14F", border = "white", breaks = 15)
abline(v = mean(coef_table$Beta3), col = "red", lwd = 2, lty = 2)
par(mfrow = c(1, 1))
```

# 6. Group Portfolios & Regression

## 6.1 Define 4 groups of stocks based on size and BtM

```{r define-portfolios}
idx_BH <- intersect(idx_big,   idx_high)   # Big / High BtM
idx_BL <- intersect(idx_big,   idx_low)    # Big / Low BtM
idx_SH <- intersect(idx_small, idx_high)   # Small / High BtM
idx_SL <- intersect(idx_small, idx_low)    # Small / Low BtM

cat(sprintf("B/H: %d Stocks | B/L: %d Stocks | S/H: %d Stocks | S/L: %d Stocks\n",
    length(idx_BH), length(idx_BL), length(idx_SH), length(idx_SL)))

r_bh <- if (length(idx_BH) > 0) rowMeans(r[, idx_BH, drop = FALSE]) else rep(NA, nrow(r))
r_bl <- if (length(idx_BL) > 0) rowMeans(r[, idx_BL, drop = FALSE]) else rep(NA, nrow(r))
r_sh <- if (length(idx_SH) > 0) rowMeans(r[, idx_SH, drop = FALSE]) else rep(NA, nrow(r))
r_sl <- if (length(idx_SL) > 0) rowMeans(r[, idx_SL, drop = FALSE]) else rep(NA, nrow(r))
```

## 6.2. Regression

```{r reg-portfolios}
reg1 <- lm(r_bh ~ F1 + F2 + F3); cat("=== B/H ===\n"); print(summary(reg1))
reg2 <- lm(r_bl ~ F1 + F2 + F3); cat("=== B/L ===\n"); print(summary(reg2))
reg3 <- lm(r_sh ~ F1 + F2 + F3); cat("=== S/H ===\n"); print(summary(reg3))
reg4 <- lm(r_sl ~ F1 + F2 + F3); cat("=== S/L ===\n"); print(summary(reg4))
```

## 6.3. VIF test

```{r vif-check}
cat("VIF test for B/H:\n"); print(vif(reg1))
cat("\nVIF test for B/L:\n"); print(vif(reg2))
cat("\nVIF test for S/H:\n"); print(vif(reg3))
cat("\nVIF test for S/L:\n"); print(vif(reg4))
```

# 7. Hypothesis Testing
## RESET-test & Breusch-Pagan test

```{r diag-tests}
run_diagnostics <- function(model, label) {
  reset <- resettest(model, type = "regressor")
  bp    <- bptest(model)
  dw    <- dwtest(model)
  bg    <- bgtest(model, order = 5)

  data.frame(
    Portfolio    = label,
    RESET_pval   = round(reset$p.value, 4),
    BP_pval      = round(bp$p.value, 4),
    DW_pval      = round(dw$p.value, 4),
    BG5_pval     = round(bg$p.value, 4),
    Form_OK      = reset$p.value > 0.05,
    Homosked_OK  = bp$p.value > 0.05,
    NoAutoCorr_OK= dw$p.value > 0.05
  )
}

diag_results <- rbind(
  run_diagnostics(reg1, "B/H"),
  run_diagnostics(reg2, "B/L"),
  run_diagnostics(reg3, "S/H"),
  run_diagnostics(reg4, "S/L")
)

kable(diag_results, caption = "Diagnostic Tests for Portfolio Regressions")
```

# 8. Estimate Risk-Premia  (Fama-MacBeth Step 2)

```{r risk-premium}
# mean-return
r_mean <- colMeans(r[, 1:N_STOCKS])

datanew <- data.frame(
  r_mean = r_mean,
  alpha  = coef_table$alpha,
  Beta1  = coef_table$Beta1,
  Beta2  = coef_table$Beta2,
  Beta3  = coef_table$Beta3
)

# r_mean ~ Beta
datanew$intercept <- 1
reg_cs <- lm(r_mean ~ intercept + Beta1 + Beta2 + Beta3 - 1, data = datanew)
summary(reg_cs)

# Estimate robust standard error (Newey-West)
coeftest(reg_cs, vcov = NeweyWest(reg_cs, lag = 5))
```

# 9. APT test

```{r apt-test}
reg_APT <- lm(r_mean ~ intercept + Beta1 + Beta2 + Beta3 - 1, data = datanew)
summary(reg_APT)

r_f_daily <- mean(r_f)
hyp_str   <- paste0("intercept = ", format(r_f_daily, scientific = FALSE, digits = 15))
cat(sprintf("\nintercept = r_f (%.8f) test:\n", r_f_daily))
cat("Hypothesis:", hyp_str, "\n")
linearHypothesis(reg_APT, hyp_str)
```

# 10. Portfolio Risk Analysis

## 10.1. Covariance Matrix of Factors and Errors

```{r cov-matrices}
VF <- cov(factor_mat)
cat("Covariance Matric of Factors:\n")
print(round(VF, 8))

resid_mat <- do.call(cbind, lapply(models, residuals))
cov_resid <- cov(resid_mat)
```

## 10.2. Risk Decomposition for Each Stock

```{r risk-decomp}
Beta_mat <- as.matrix(coef_table[, c("Beta1", "Beta2", "Beta3")])

risk_table <- data.frame(
  Stock             = rownames(coef_table),
  Systematic_Risk   = numeric(N_STOCKS),
  Idiosyncratic_Risk= numeric(N_STOCKS),
  Total_Risk        = numeric(N_STOCKS),
  R_squared         = sapply(models, function(m) summary(m)$r.squared)
)

for (i in 1:N_STOCKS) {
  # Systematic Risk: β' Σ_F β 
  risk_table$Systematic_Risk[i]    <- as.numeric(Beta_mat[i, ] %*% VF %*% Beta_mat[i, ])
  risk_table$Idiosyncratic_Risk[i] <- var(resid(models[[i]]))
  risk_table$Total_Risk[i]         <- risk_table$Systematic_Risk[i] + risk_table$Idiosyncratic_Risk[i]
}

```

```{r plot-risk}
# Proportion of systematic risk to total risk Plot
risk_table$Sys_Ratio <- risk_table$Systematic_Risk / risk_table$Total_Risk

ggplot(risk_table, aes(x = reorder(Stock, Sys_Ratio), y = Sys_Ratio)) +
  geom_col(fill = "#4E79A7", alpha = 0.85) +
  geom_hline(yintercept = mean(risk_table$Sys_Ratio), color = "red",
             linetype = "dashed", linewidth = 0.8) +
  coord_flip() +
  labs(title    = "Proportion of Systematic Risk to Total Risk for Each Stock",
       subtitle = paste0("Red line: Mean = ",
                         round(mean(risk_table$Sys_Ratio) * 100, 1), "%"),
       x = "Stock", y = "Proportion of Systematic Risk") +
  theme_minimal(base_size = 10)
```

## 10.3. Equal-Weight Portfolio Risk

```{r portfolio-risk}
n  <- N_STOCKS
w  <- matrix(rep(1/n, n), nrow = 1)           # Equal-weight
w_vec <- as.numeric(w)

Beta_p  <- w %*% Beta_mat                      # Portfolio Beta (1x3)
Sys_p   <- as.numeric(Beta_p %*% VF %*% t(Beta_p))    # Portfolio Systematic Risk
Idio_p  <- sum(w_vec^2 * risk_table$Idiosyncratic_Risk)   # Portfolio Idiosyncratic Risk (assuming no correlation between residuals)
Total_p <- Sys_p + Idio_p

cat("=== Equal-Weight Portfolio (1/50) ===\n")
cat(sprintf("Portfolio Beta:        %.5f  %.5f  %.5f\n", Beta_p[1], Beta_p[2], Beta_p[3]))
cat(sprintf("Systematic Risk:      %.8f\n", Sys_p))
cat(sprintf("Idiosyncratic Risk:  %.8f\n", Idio_p))
cat(sprintf("Total Risk:          %.8f\n", Total_p))
cat(sprintf("Rate of Diversification:    %.2f%%\n", (1 - Sys_p/Total_p)*100))
```
